A neutrality index for quantitative traits from within and between variations leveraging sequences evolution

### 2023-02-03 manuscript/main.tex
Intro/discussion

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires to determine which model of evolution is supported from the observed variation of traits and sequences.
This variation of traits can be observed at different scales, at the individual level across different development stages, at the population level across different individuals, at the species level across different populations, and finally at the phylogenetic level across different species.
All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
At the species level, by contrasting variation within and between different populations, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes\cite{whitehead_neutral_2006, gilad_natural_2006, gilad_expression_2006}.
However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
For a given population, by contrasting to the divergence to a close species, studies have confirms that changes in gene expression accumulate linearly differences with time from divergence\cite{khaitovich_neutral_2004}.
% This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
% However, it has not yet been generalized for any number of species related by their phylogenetic tree.
% However, this divergence in the level of expression stabilizes over a long distance: the difference between human and chicken (having diverged 300 million years) is similar to that between human and opossum (160 million years).
% This observation suggests that the conservation of essential functions of organs could define a limit for the divergence of transcriptomes.

% Phylogenetics (mean trait evolution)
At the phylogenetic scale, by determining how much the mean trait value (taken as average across different individual for a given species) is changing across different species, the mode of selection for a quantitative trait can also be tested while accounting for the underlying relationships between species\cite{felsenstein_phylogenies_1985}.
Typically, the mean trait value is modelled as a Brownian process, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
Alternatively, a deviation from the Brownian process, typically an Ornstein-Uhlenbeck process is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
However, studies have shown that comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
First, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
Indeed, a trait under selection for which the optimal trait is evolving as a Brownian process will not deviate from a Brownian process\cite{hansen_translating_1996}.
Secondly and contrarily, even under a neutral trait the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling   artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.

% At the frontier Phylogenetics/Population-genetics
At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Altogether, there is a push toward integrating within and between species variation into a unified framework.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait, leveraging data within and between species, normalizing the trait evolution with both genetic polymorphism within species and divergence between species.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, without prior knowledge, estimating a statistic called $\RatePhy / \RatePop$.
Technically, $\RatePhy / \RatePop$ is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.

@@

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
Quantitative genetics formalism has related both within and between species variation to the underlying mutational input for a neutral trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

@@

\section{Discussion}\label{sec:discussion}
% Consequences of our neutrality index
% Use case for gene expression level
 Distinguishing adaptive changes driven by positive selection from neutral divergence resulting from mutation and genetic drift is critical for understanding the evolution of a quantitative trait.
At the phylogenetic scale, tests of neutrality for a quantitative trait have been developed, modelling the underlying process generating the change in mean trait value across different species at the phylogenetic scale\cite{fay_evaluating_2008}.
These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
Our diversity index, by leveraging trait variation at different scale and normalizing by divergence and polymorphism, we can set a threshold for neutrality and thus is able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
To do so, our neutrality index leverage sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
We argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.

% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
More theoretically, our development is based on several assumption that could be relaxed such as constant genetic architecture and independence of traits.
First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
Moreover, even though the neutrality index can be applied for has many traits as available,  each trait is so far assumed to be independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003,  lartillot_phylogenetic_2011,lartillot_joint_2012}.

>>>>

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires to determine which mode of selection is supported from the observed variation of traits and sequences.
This variation of traits can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.
At the population level across individuals, Genome Wide Association Studies (GWAS) in humans have revealed that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection\cite{simons_population_2018, sella_thinking_2019}.
At the species level, by contrasting variation within and between different populations, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes\cite{whitehead_neutral_2006, gilad_natural_2006, gilad_expression_2006}.
However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated\cite{signor_evolution_2018, price_detecting_2022}.
For a given population, by contrasting to the divergence to a close species, studies have confirms that changes in gene expression accumulate linearly differences with time from divergence\cite{khaitovich_neutral_2004}.
However, studies including variation across individuals and population typically include species divergence to a close species in a pairwise comparison, and do not integrate the changes across the whole phylogenetic tree.
% This relationship had been proposed to describe the evolutionary regime of a trait for a pair of species\cite{lande_genetic_1980, turelli_heritable_1984}.
% However, it has not yet been generalized for any number of species related by their phylogenetic tree.
% However, this divergence in the level of expression stabilizes over a long distance: the difference between human and chicken (having diverged 300 million years) is similar to that between human and opossum (160 million years).
% This observation suggests that the conservation of essential functions of organs could define a limit for the divergence of transcriptomes.

% Phylogenetics (mean trait evolution)
Contrarily, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale\cite{felsenstein_phylogenies_1985}.
Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a variance proportional to the length of the branch\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
Alternatively, a deviation from the Brownian process, typically an Ornstein-Uhlenbeck process is interpreted as a signature of stabilizing selection\cite{catalan_drift_2019}.
However, studies have shown that comparative approaches are subject to different biases\cite{harmon_phylogenetic_2018}.
First, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
Indeed, a trait under selection for which the optimal trait value is evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral\cite{hansen_translating_1996}.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling   artifacts\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Indeed, in such framework, the variation within species is discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
Altogether, the classical comparative framework allows tests for neutrality at the phylogenetic scale but discards the variance in traits between individuals, leaving out valuable information.

% At the frontier Phylogenetics/Population-genetics
At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within populations is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait variation at both the phylogenetic and population scale.
Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales.

@@

To determine whether a trait is neutrally evolving or under selection, we reviewed the theoretical equations of trait evolution at the population and phylogenetic scale.
The underlying mutational input for a neutral trait is formally related to both within and between species variation of the trait.
More specifically, for a heritable trait under a neutral regime, the genetic variance of the trait ($\VarGenetic$) normalized by genetic diversity of any neutral genomic region ($\pi$) is a proxy for the mutational variance ($\RatePop$).
At the phylogenetic level, along a specific lineage, the mean value for the trait evolves as a Brownian process, and the variance in this mean value after a time $\Time$ is called $\VarPhy$.
This variance can also be normalized by the genetic distance of any neutral genomic region ($d$), allowing us to obtain another proxy for the mutational variance ($\RatePhy$).
Altogether, we define the neutrality index as $\RatePhy/\RatePop$, which equals to $1$ for a neutral trait, suggesting that traits for which this relationship is not verified are putatively under a selective regime.
Under a depleted variation between species, meaning $\RatePhy < \RatePop$, the trait is putatively under stabilizing, forcing the mean value to be homogeneous between species.
In contrast, under an inflated variation between species, meaning $\RatePhy > \RatePop$, the trait is putatively under diversifying selection, or a moving optimum such that the mean value is heterogeneous in the different species.

@@

\section{Discussion}\label{sec:discussion}

In this study, we proposed a neutrality index for a quantitative trait.
At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
Similarly, at the population scale, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
Our neutrality index for a trait is the ratio of between species variation to within species variation.
Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
Contrarily, we argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.

At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($h^2$).

% Assumption that could be relaxed: constant genetic architecture and multidimensional generalization
More theoretically, our development is based on several assumption that could be relaxed.
First, $\RateMut$ is assumed to be constant across the phylogeny, while it is determined by the underlying genetic architecture of a trait, which can be variable among individuals and species.
Having a constant $\RateMut$ across the phylogeny is an assumption that can theoretically by relaxed by adding a time component\cite{arnold_understanding_2008, hohenlohe_mipod_2008}, where $\RateMut(t)$ could jointly be estimated along the phylogeny\cite{kostikova_bridging_2016, gaboriau_multiplatform_2020}.
Moreover, even though the neutrality index can be applied for has many traits as available, each trait is so far assumed to be independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework\cite{huelsenbeck_detecting_2003,  lartillot_phylogenetic_2011,lartillot_joint_2012}.
Altogether, we argue that our neutrality index is a promising statistical procedure allowing to determine the evolutionary regime of a quantitative trait, through articulation of the variance between and within populations in a unified framework.
